---
title: "Luke (2004) Multilevel Modeling in lme4"
Author: "Angus Hughes"
output: md_document
---

#Background

> The aim of this analysis is to reproduce the results from Douglas A. Luke's "Multilevel Modeling" (2004) by Sage University Press using lme4 and the tidyverse (ggplot, dplyr) from the original dataset.

The dataset is taken from a tobacco control policy study, with the main goal to identify important influences on voting on tobacco-related legislation by members of Congress from 1997 to 2000 (Luke & Krauss, 2004). 

The dependent variable is Voting % ("VotePct"), the percentage of time that a senator or representative voted in a "pro-tobacco" direction during those four years. A "No" vote was considered to be a "pro-tobacco" vote because the tobacco industry opposed this legislation. Voting % will be calculated for each member of Congress by adding up the total number of times that he or she voted in a pro-industry direction, and dividing by the total number of tobacco-related bills that he or she voted on. The variable can range from 0.0 (never voted protobacco) to 1.O (always voted pro-tobacco). 

Independent variables for each politician include: 

*   *Political Party*. Past research has shown that political party is an important predictor of voting pattern - Republicans tend to vote more often in the pro-tobacco industry direction. 
*   *Money*. The amount of money that the member of Congress received from tobacco industry political action committees (PACs). Our hypothesis is that the more PAC money a legislator receives, the more often that person will vote pro-tobacco.

We also include two higher level independent variables:

*   *State*. which state each member of Congress represents.
*   *State tobacco farm economy*. We measure this with Acres, the number of harvested acres of tobacco in 1999, in thousands of acres.

#Data Preparation

##Tidy the Dataset

Need to calculate the voting percentage to achieve same dependent variable as Luke (2004).

```{r tidy dataset, warning=FALSE,message=FALSE}
library(tidyverse)
load("../Data/tobvote.rda")

# Load tobvote from data file
tobacco <- tobvote %>% 
  select(lastname, firstnam, state, house, pactotal, party, votedpro, votenum, acres) %>% #extract variables of interest
  group_by(state, lastname, house)%>% #Arrange in state and last name order
  mutate(vote = (as.numeric(votedpro)-1), #covert voting behaviour to binary numeric data
                   VotePct = round((sum(vote)/votenum),2)) %>% #calculate percentage of pro-tobacco votes
  select(-vote, -votedpro) %>% #remove nuisance variables
  unique() #reduce data frame to one row per representative
```

##Tables 2.2 & 2.3 "Structure of Level-1 Tobacco Data Set" and "Structure of Level-2 Tobacco Data Set"

```{r view data}
tobacco
```

Note there is a discrepancy here - the text had $N = 527$. We appear to be missing one congress person from California. Otherwise the data match exactly.

#Visualisations

##Figure 2.2 "Average Pro-Tobacco Vote Percentage by Congress Members - 1997 to 2000"

```{r fifty state map, warning=FALSE,message=FALSE,error=FALSE}
# Need to install "fiftystater" package for Alaska and Hawaii
if(!require(fiftystater)){
  if(!require(devtools)){install.packages("devtools")}
  else devtools::install_github("wmurphyrd/fiftystater");
  install.packages("mapproj")}
library("fiftystater")
library("mapproj")

vote_by_state <- tobacco %>% group_by(state) %>% 
  summarise(Avg = mean(VotePct)) %>% 
  mutate(statename = tolower(state.name[match(state, state.abb)]))

state_plot <- ggplot(vote_by_state, aes(map_id = statename)) + 
  # map points to the fifty_states shape data
  geom_map(aes(fill = Avg), map = fifty_states)+ 
  expand_limits(x = fifty_states$long, y = fifty_states$lat) +
  geom_text(aes(state.center$x, state.center$y, label = state.abb), size = 3) + 
  coord_map() +
  scale_fill_gradient2(name = "Avg Pro-Tobacco Vote", high = "grey10") +
  scale_x_continuous(breaks = NULL) + 
  scale_y_continuous(breaks = NULL) +
  labs(x = "", y = "") +
  theme(legend.position = "bottom", 
        panel.background = element_blank()) 
state_plot
```

Alaska and Hawaii labels don't play nice. Will come back to this. Otherwise the visualisations are roughly accurate to the text.
https://cran.r-project.org/web/packages/fiftystater/vignettes/fiftystater.html

##Figure 2.3 "Relationship of Money and Voting % for 5 Large States"

```{r 5 state money vote, warning=FALSE,message=FALSE,error=FALSE}
five_states_plot <- tobacco %>% 
  filter(state == "CA" |state ==  "FL" |state ==  "IL" |state ==  "NY" |state ==  "TX") %>% 
  ggplot(aes(x = pactotal, y = VotePct, shape = state, linetype = state)) + 
    geom_point() + ylim(0,1) + xlim(0,80) + labs(x = "PAC Contributions ($K)", y = "Pro-Tobacco Voting (%)") +
    stat_smooth(method = "lm", se =FALSE, fullrange = T, color = "black", size=.3) + theme_classic() 
five_states_plot
```

Graph mirrors that of the book very closely.

##Figure 2.4 "OLS Fits of Voting % and Money for 50 States"

```{r OLS fits for all states}
#This may take some time to run
ols_states_plot <- ggplot(tobacco, aes(x = pactotal, y = VotePct)) + 
  geom_point(shape = 1, size = 2, alpha = .8) +
  stat_smooth(method = "lm", se =FALSE, fullrange = T, color = "black", size=.3) + 
  facet_wrap( ~ state) + theme_bw() + 
  theme(panel.spacing = unit(0, "lines"), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  labs(x = "PAC Money", y = "Vote %") +
  ylim(0,1) + xlim(0,100) 
ols_states_plot
```

Aesthetics aren't exactly the same, but OLS fits are more or less identical and now in alphabetical order. Slopes and intercepts clearly vary by state.

#Model Fitting

This is implemented in lme4, rather than nmle. Slight differences in estimation are evident, but results are similar.

##Table 2.5 "Partial R Output for the Null Model"

```{r Null Model, warning=FALSE,error=FALSE,message=FALSE}
library(lme4)
library(lmerTest)
null_fit <- lmer(VotePct ~ 1 + (1 | state), tobacco, REML = F)
summary(null_fit)
```

Model fits differ somewhat, but result is extremely close.

##Table 2.7 "Partial R Output for Model 1"

```{r Model 1, warning=FALSE,error=FALSE,message=FALSE}
model1_fit <- lmer(VotePct ~ 1+ pactotal + party + (pactotal + party | state), tobacco, REML = F)
summary(model1_fit)
```

Results are similar.

##Table 2.9a "Parameter Estimates and Model Fit for Three Models"

```{r Model 2, warning=FALSE,error=FALSE,message=FALSE}
model2_fit <- lmer(VotePct ~ 1 + pactotal + party + acres + (pactotal + party | state), tobacco, REML = F)
summary(model2_fit)
```

Fixed effects estimates and variance components are roughly similar. lmerTest provides significance tests for the former, which resemble published results, with the exception of 'acres' variable which is here non-significant. Neither lmerTest nor lme4 provide significance tests for variance components.

Difficult to say why differences emerge - may be due to software (HLM providing the published estimates?) implementing a different form of test. Need to double check this.

##Table 2.9b "Parameter Estimates and Model Fit for Three Models"

```{r Model 3, warning=FALSE,error=FALSE,message=FALSE}
model3_fit <- lmer(VotePct ~ 1 + pactotal + party + acres + pactotal*acres + party*acres + (pactotal + party | state), tobacco, REML = F)
summary(model3_fit)
```

Again, t-value and significance levels differ from published results. Unclear why this may be, although lme4 provides a warning about rescaling the variable (acres ranges from many observations of zero to >1,000).

#More Visualisations

##Figure 2.5 "Predicted Pro-Tobacco Voting Percent for Democrats by Amount of Tobacco Acreage"

```{r Plot Predicted by Acreage}
#NOT RUN - COME BACK TO THIS
#model3.re <- ranef(model3.fit) %>% .$state
#tobacco %>% group_by(state) %>% 
#  summarise(mean=mean(acres)) %>% 
#  mutate(quantile = cut(mean, breaks = c(0,0.1,200,500), labels = F, include.lowest = T))
```

Currently not implemented - still need to determine how cut-offs between no acreage, moderate acreage and high acreage were determined by Luke (2004).

##Figure 2.6 "Boxplots of Within-State residuals for Model 3"

```{r Residual Plot}
residual_plot <- ggplot(model3_fit, aes(y = residuals(model3_fit), x = reorder(state, desc(state)))) +
  geom_boxplot(outlier.shape = 1, size = .4) + coord_flip() + geom_hline(yintercept = 0) +
  theme_classic() + labs(x = "State", y = "Residuals")
residual_plot
```

Now in alphabetical order. Visualisation is approximately the same.

##Figure 2.7 "Scatterplot of Standardised Residuals by Fitted Values for Model 3 by Party"

```{r Standardised Residual Plot}
std_residual_plot <- ggplot(model3_fit, aes(.fitted, .resid)) + 
  geom_point(shape = 1) +
  facet_wrap(~ party) + 
  theme_bw() + theme(panel.grid.minor = element_blank()) +
  geom_text(aes(label=ifelse(.resid>0.5 | .resid <=-0.5,as.character(state),'')),hjust=0,vjust=0) +
  labs(x="Fitted Values",y="Standardised Values")
std_residual_plot
```

Visualisation is approximately the same.

##Figure 2.8 "Normal QQ-Plot of Residuals of Model 3 by Party"

```{r}
qplot_party <- ggplot(model3_fit, aes(sample = residuals(model3_fit, scaled = FALSE))) + 
  geom_point(stat = "qq", shape = 1) + coord_flip() + facet_wrap(~party) +
  theme_bw() + theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank()) +
  labs(x="Quantiles of Standard Normal", y="Residuals")
qplot_party
```

Visually similar, although some observations amongst democrats at the low quantiles appear further.

##Figure 2.9 "Normal QQ-Plot of Random Effects of Model 3"

```{r}
nested_re <- as.data.frame(ranef(model3_fit))

nested_re %>% 
  ggplot(aes(sample = condval)) +
  stat_qq(aes(group = 1), alpha = 0.5) +
  coord_flip() + theme_bw() +
  facet_wrap(~term,  scales = "free") +
  labs(y = "Quantiles of Standard Normal", x = "Random Effects")
```


# Figure 2.10 "Scatterplot Matrix"

```{r}
nested_re %>% 
  select(-condsd) %>% 
  spread(term, condval, convert = TRUE) %>% 
  pairs(~ `(Intercept)` + pactotal + partyRepublican, 
                         data = .)
```

