---
title: "Luke (2004) Multilevel Modeling in lme4"
Author: "Angus Hughes"
output: pdf_document
---

#Background

> The aim of this analysis is to reproduce the results from Douglas A. Luke's "Multilevel Modeling" (2004), Sage University Press, using lme4 and the tidyverse from the original dataset.

The dataset is taken from a tobacco control policy study, with the main goal to identify important influences on voting on tobacco-related legislation by members of Congress from 1997 to 2000 (Luke & Krauss, 2004). 

The dependent variable is Voting % ("VotePct"), the percentage of time that a senator or representative voted in a "pro-tobacco" direction during those four years. A "No" vote was considered to be a "pro-tobacco" vote because the tobacco industry opposed this legislation. Voting % will be calculated for each member of Congress by adding up the total number of times that he or she voted in a pro-industry direction, and dividing by the total number of tobacco-related bills that he or she voted on. The variable can range from 0.0 (never voted protobacco) to 1.O (always voted pro-tobacco). 

Independent variables for each politician include: 

*   *Political Party*. Past research has shown that political party is an important predictor of voting pattern - Republicans tend to vote more often in the pro-tobacco industry direction. 
*   *Money*. The amount of money that the member of Congress received from tobacco industry political action committees (PACs). Our hypothesis is that the more PAC money a legislator receives, the more often that person will vote pro-tobacco.

We also include two higher level independent variables:

*   *State*. which state each member of Congress represents.
*   *State tobacco farm economy*. We measure this with Acres, the number of harvested acres of tobacco in 1999, in thousands of acres.

#Data Preparation

##Tidy the Dataset

Need to calculate the voting percentage to achieve same dependent variable as Luke (2004).

```{r tidy dataset, warning=FALSE,message=FALSE}
library(tidyverse)
options(scipen = 99)
load("../Data/tobvote.rda")

# Load tobvote from data file
tobacco <- tobvote %>% 
  select(lastname, firstnam, state, house, pactotal, party, votedpro, votenum, acres) %>% 
  group_by(state, lastname, firstnam, house)%>% 
  mutate(vote = (as.numeric(votedpro)-1),      # Convert vote to binary numeric 
         VotePct = (sum(vote)/votenum)) %>%  # Calculate % pro-tobacco votes
  select(-vote, -votedpro) %>% 
  unique()                         # Reduce data to one row per representative
```

##Tables 2.2 & 2.3 "Structure of Level-1 Tobacco Data Set" and "Structure of Level-2 Tobacco Data Set"

```{r view data}
tobacco %>% mutate(VotePct = round(VotePct, 2))
```

Note there is a discrepancy here - the text had $N = 527$. We appear to be missing one congress person from California. Otherwise the data match exactly.

#Visualisations

##Figure 2.2 "Average Pro-Tobacco Vote Percentage by Congress Members - 1997 to 2000"

```{r fifty state map, warning=FALSE,message=FALSE,error=FALSE}
# Need to install "fiftystater" package for Alaska and Hawaii
if(!require(fiftystater)){
  if(!require(devtools)){install.packages("devtools")}
  else devtools::install_github("wmurphyrd/fiftystater");
  install.packages("mapproj")}
library("fiftystater")
library("mapproj")

vote_by_state <- tobacco %>% 
  group_by(state) %>% 
  summarise(Avg = mean(VotePct)) %>% 
  mutate(statename = tolower(state.name[match(state, state.abb)]))

state_plot <- ggplot(vote_by_state, aes(map_id = statename)) + 
  # map points to the fifty_states shape data
  geom_map(aes(fill = Avg), map = fifty_states)+ 
  expand_limits(x = fifty_states$long, y = fifty_states$lat) +
  geom_text(aes(state.center$x, state.center$y, label = state.abb), size = 3) + 
  coord_map() +
  scale_fill_gradient2(name = "Avg Pro-Tobacco Vote", high = "grey10") +
  scale_x_continuous(breaks = NULL) + 
  scale_y_continuous(breaks = NULL) +
  labs(x = "", y = "") +
  theme(legend.position = "bottom", 
        panel.background = element_blank()) 

state_plot
```

Alaska and Hawaii labels don't play nice. Will come back to this. Otherwise the visualisations are roughly accurate to the text.
https://cran.r-project.org/web/packages/fiftystater/vignettes/fiftystater.html

##Figure 2.3 "Relationship of Money and Voting % for 5 Large States"

```{r 5 state money vote, warning=FALSE,message=FALSE,error=FALSE}
five_states_plot <- tobacco %>% 
  filter(state == "CA" |state ==  "FL" |state ==  "IL" |state ==  "NY" |state ==  "TX") %>% 
  ggplot(aes(x = pactotal, y = VotePct, shape = state, linetype = state)) + 
  geom_point() + 
  stat_smooth(method = "lm", se =FALSE, fullrange = T, color = "black", size=.3) +
  ylim(0,1) + xlim(0,80) + 
  labs(x = "PAC Contributions ($K)", y = "Pro-Tobacco Voting (%)") +
  theme_classic() 

five_states_plot
```

Graph mirrors that of the book very closely.

##Figure 2.4 "OLS Fits of Voting % and Money for 50 States"

```{r OLS fits for all states, echo = TRUE, message = FALSE, warning = FALSE, fig.align="center", fig.cap="Slopes and intercepts clearly appear to vary by state."}
#This may take some time to run
ols_states_plot <- ggplot(tobacco, aes(x = pactotal, y = VotePct)) + 
  geom_point(shape = 1, size = 1.25, alpha = .8) +
  stat_smooth(method = "lm", se =FALSE, fullrange = T, 
              color = "black", size=.3, alpha = .75) + 
  facet_wrap( ~ state) + 
  theme_bw() + 
  theme(panel.spacing = unit(0, "lines"), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  labs(x = "PAC Money", y = "Vote %") +
  ylim(0,1) + xlim(0,100) 

ols_states_plot
```

Aesthetics aren't exactly the same, but OLS fits are more or less identical and now in alphabetical order.

#Model Fitting

This is implemented in lme4, rather than nmle. Slight differences in estimation are evident, but results are similar.

##Table 2.5 "Partial R Output for the Null Model"

```{r Null Model, warning=FALSE,error=FALSE,message=FALSE}
library(lme4)
library(lmerTest)

null_fit <- lmer(VotePct ~ 1 + (1 | state), 
                 data = tobacco, REML = F)
summary(null_fit)
```

Model fits differ somewhat, but result is extremely close.

##Table 2.7 "Partial R Output for Model 1"

```{r Model 1, warning=FALSE,error=FALSE,message=FALSE}
model1_fit <- lmer(VotePct ~ pactotal + party +   # Fixed effects
                     (pactotal + party | state),  # Random effects
                   data = tobacco, REML = F)

summary(model1_fit)
```

Results are similar.

##Table 2.9a "Parameter Estimates and Model Fit for Three Models"

```{r Model 2, warning=FALSE,error=FALSE,message=FALSE}
model2_fit <- lmer(VotePct ~ pactotal + party + acres + 
                     (pactotal + party | state), 
                   data = tobacco, REML = F)

summary(model2_fit)
```

Fixed effects estimates and variance components are roughly similar. lmerTest provides significance tests for the former, which resemble published results, with the exception of 'acres' variable which is here non-significant. Neither lmerTest nor lme4 provide significance tests for variance components.

##Table 2.9b "Parameter Estimates and Model Fit for Three Models"

```{r Model 3, warning=FALSE,error=FALSE,message=FALSE}
model3_fit <- lmer(VotePct ~ pactotal + party + acres + 
                     pactotal*acres + party*acres +      # Interactions
                     (pactotal + party | state), 
                   data = tobacco, REML = F)

summary(model3_fit)
```

Again, t-value and significance levels differ from published results. Unclear why this may be, although lme4 provides a warning about rescaling the variable (acres ranges from many observations of zero to >1,000).

##Figure 2.5 "Predicted Pro-Tobacco Voting Percent for Democrats by Amount of Tobacco Acreage"

For ease of plotting, I'll use the `emmeans` package. This also helpfully provides the confidence bands, improving upon the original.

```{r Plot Predicted by Acreage}
library(emmeans)

acreages <- c("No Acreage", "Moderate Acreage", "High Acreage")
demo_data <- emmeans(model3_fit, c("party", "pactotal", "acres"), 
                     at = list(party = "Democrat", 
                               pactotal = seq.int(0, 100, 10),
                               acres = c(0, 33, 200)),
                     type = "response") %>% as.data.frame %>% 
  mutate(acres = factor(acres, labels = acreages))

demo_plot <- ggplot(demo_data, aes(y = emmean, x = pactotal, group = acres)) +
  geom_line(aes(linetype = acres)) +
  geom_ribbon(aes(ymin = lower.CL, ymax = upper.CL), alpha = .05) +
  theme_bw() +
  coord_cartesian(ylim = c(0,.96), xlim = c(2, 98)) +
  labs(y = "Predicted Voting - Democrats (%)",
       x = "PAC Contributions ($K)") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = c(0.85, 0.15),
        legend.title = element_blank())

demo_plot
```

Results reproduced with standard error represented.

#Model Diagnostics

##Figure 2.6 "Boxplots of Within-State residuals for Model 3"

```{r Residual Plot}
residual_plot <- ggplot(model3_fit, aes(y = residuals(model3_fit), x = reorder(state, desc(state)))) +
  geom_boxplot(outlier.shape = 1, size = .4) + 
  coord_flip() + 
  geom_hline(yintercept = 0) +
  theme_classic() + labs(x = "State", y = "Residuals")

residual_plot
```

Now in alphabetical order. Visualisation is approximately the same.

##Figure 2.7 "Scatterplot of Standardised Residuals by Fitted Values for Model 3 by Party"

```{r Standardised Residual Plot}
std_residual_plot <- ggplot(model3_fit, aes(.fitted, .resid)) + 
  geom_point(shape = 1) +
  facet_wrap(~ party) + 
  theme_bw() + theme(panel.grid.minor = element_blank()) +
  geom_text(aes(label = ifelse(.resid > 0.5 | .resid <= -0.5, # Label if extreme
                               as.character(state), '')),     
            hjust = 0, vjust = 0) +
  labs(x = "Fitted Values", y = "Standardised Values")

std_residual_plot
```

Visualisation is approximately the same.

##Figure 2.8 "Normal QQ-Plot of Residuals of Model 3 by Party"

```{r Normal QQ plot of residuals}
qplot_party <- ggplot(model3_fit, aes(sample = residuals(model3_fit, scaled = FALSE))) + 
  geom_point(stat = "qq", shape = 1) + 
  coord_flip() + 
  facet_wrap( ~ party) +
  theme_bw() + 
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank()) +
  labs(x = "Quantiles of Standard Normal", y = "Residuals")

qplot_party
```

Visually similar, although some observations amongst democrats at the low quantiles appear further.

## Figure 2.9 "Normal QQ-Plot of Random Effects of Model 3"

```{r Normal RE QQ Plot}
random_effects <- as.data.frame(ranef(model3_fit))

qplot_re <- random_effects %>% 
  ggplot(aes(sample = condval)) + 
  stat_qq(aes(group = 1), alpha = 0.5) +
  coord_flip() + theme_bw() +
  facet_wrap(~term,  scales = "free") +
  labs(x = "Quantiles of Standard Normal", y = "Random Effects")

qplot_re
```

## Figure 2.10 "Scatterplot Matrix"

This is accomplished with a simple `pairs()` call in the book. We need to use `ggpairs()` from the "GGally" package to replicate the same in ggplot2.

```{r Scatterplot matrix, message=FALSE, warning=FALSE}
library(GGally)

scatterplot_re <- random_effects %>% 
  select(-condsd) %>% 
  spread(term, condval, convert = TRUE) %>% 
  GGally::ggpairs(columns = c("(Intercept)", "pactotal", "partyRepublican")) + 
  theme_bw()

scatterplot_re
```

# Best Linear Unbiased Estimators

## Table 2.11

Luke here actually computes combined parameter estimates of the Intercept, pactotal (Money), and party by state *incorporating* the true state acreage value and the interaction term with acreage with the latter two variables.

The described "Empirical Bayes" estimates are obtained via `coef()` in lme4. These go by various different terms, such as best linear unbiased estimators (BLUEs) or predictors (BLUPs), conditional mode of parameters by Doug Bates; in the text, Luke also refers to them as "shrinkage estimates". 

We adjust these BLUEs with the fixed effects estimates for interaction terms and acreage of the state.

```{r BLUE Table}
# Prepare the data
state_data <- tobacco %>% 
  group_by(state) %>% 
  summarise(acres = first(acres),
            legislators = n())

# Best Linear Unbiased Estimates
fixed_effects <- fixef(model3_fit)
blues <- coef(model3_fit)$state
# Change "(Intercept)" to "Intercept" for convenience
colnames(blues)[1] <- names(fixed_effects)[1] <- "Intercept"

# Custom function to create the EB estimates
adj_est <-  function(a){
  term <- ifelse(a == "Intercept", "acres", paste(a, ":acres", sep = ""))
  blues[[a]] + (state_data[["acres"]] * fixed_effects[[term]])
}

# Calculate EB estimates
eb_estimates <- blues %>% 
  transmute(State = state_data[["state"]],
            Legislators = state_data[["legislators"]],
            Intercept = adj_est("Intercept"),
            Party = adj_est("partyRepublican"),
            Money = adj_est("pactotal"),
            Acres = state_data[["acres"]])

eb_estimates
```

## Figure 2.11 "Bayesian Estimated Regression Lines for Model 3 by Party"

I plot this by calculating the estimated vote with no money (minimum), and the estimated vote with $120k worth of PAC money (maximum). These form the limits on the y axis of each regression line by state and party. I then use `geom_segment` to link these.

```{r EB party slope plot}
eb_predictions <- eb_estimates %>% 
  transmute(Dmin = Intercept + Money,
            Dmax = Intercept + (Money * 120),
            Rmin = Intercept + Party + Money,
            Rmax = Intercept + Party + (Money * 120),
            state = state_data[["state"]])

eb_party_plot <- eb_predictions %>% ggplot() + 
  geom_segment(aes(x = 0, xend = 120, y = Dmin, yend = Dmax), 
               linetype = "solid", alpha = .75) +
  geom_segment(aes(x = 0, xend = 120, y = Rmin, yend = Rmax), 
               linetype = "longdash", alpha = .75) +
  theme_minimal() +
  coord_cartesian(ylim = c(0,.96), xlim = c(5, 115)) +
  labs(y = "Predicted Pro-Tobacco Voting (%)",
       x = "PAC Contributions ($K)")

eb_party_plot
```

# Figure 2.12

Something interesting is happening here. The text states that this is a plot where "party is ignored". The trouble is that the EB estimates we have will give different intercepts and slopes depending upon the party. If we arbitrarily choose one, we bias the plot. If we take the mean across parties, we bias again because some states lean Republican, or Democrat. 

The solution here is to calculated a weighted mean of the EB estimates by the proportion of Republican legislators. This is done by calculating the ratio of R/D for each state, then using these as weights to average the minimum and maximum predicted values calculated above for each state, using a call to `sapply`.

```{r BLUE by state}
# Find ratio of Democrat to Republicans
party_ratio <- tobacco %>% 
  group_by(state, party) %>% 
  summarise(legislators = n()) %>% 
  mutate(Pct_Republican = legislators / sum(legislators)) %>% 
  complete(party, fill = list(Pct_Republican = 0)) %>% 
  ungroup() %>% select(-legislators) %>% 
  spread(key = party, value = Pct_Republican)

# Custom function to find weighted EB estimates
eb_weighted <- function(D, R){
  sapply(1:50, function(x){
    weighted.mean(x = c(eb_predictions[[D]][x], eb_predictions[[R]][x]), 
                  w = c(party_ratio[["Democrat"]][x], party_ratio[["Republican"]][x]))
  })
}
eb_by_state <- data.frame(state = state_data[["state"]],
                          Avgmin = eb_weighted("Dmin", "Rmin"),
                          Avgmax = eb_weighted("Dmax", "Rmax"))

# Now plot
eb_state_plot <- tobacco %>% 
  ggplot(aes(x = pactotal, y = VotePct)) + 
  geom_point(shape = 1, size = 1.25, alpha = .8) + 
  coord_cartesian(ylim = c(0,1), xlim = c(0, 100)) +
  facet_wrap( ~ state) +
  geom_segment(data = eb_by_state, 
               aes(x = 0, xend = 120, 
                   y = Avgmin, yend = Avgmax)) +
  theme_bw() + 
  theme(panel.spacing = unit(0, "lines"), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  labs(y = "Vote %", x = "PAC Money")

eb_state_plot
```

Results look fairly close to the published plot.

#Centering

# Table 2.12

```{r grand and state centered, message=FALSE, warning=FALSE}
mean_pac <- mean(tobacco$pactotal)
tobacco <- tobacco %>% 
  group_by(state) %>% 
  mutate(pactotal_gc = pactotal - mean_pac,
         pactotal_sc = pactotal - mean(pactotal))

model3_gc <- lmer(VotePct ~ pactotal_gc + party + acres + 
                    pactotal_gc*acres + party*acres + 
                    (pactotal_gc + party | state), 
                   data = tobacco, REML = F)
model3_sc <- lmer(VotePct ~ pactotal_sc + party + acres + 
                    pactotal_sc*acres + party*acres + 
                    (pactotal_sc + party | state), 
                   data = tobacco, REML = F)

summary(model3_gc)
```

#Generalised Linear Models

## Table 3.2

```{r first glm fit, message=FALSE, warning=FALSE}
tobacco_votes <- tobvote %>% 
  select(lastname, firstnam, state, house, pactotal, party, votedpro, votenum, acres) %>%
  mutate(vote = (as.numeric(votedpro)-1)) %>% 
  group_by(state) %>% 
  mutate(legislator = paste(firstnam, lastname))

model4_fit <- glmer(vote ~ 1 + party + pactotal + (1 | legislator), 
                   data = tobacco_votes, 
                   family = binomial)

summary(model4_fit)
```

We also need the dispersion index, not normally calculated by lme4.

```{r Dispersion index function}
# This code is nicked from the blmeco package
dispersion_glmer <- function(modelglmer){
  # computing  estimated scale  ( binomial model) following  D. Bates:
  # That quantity is the square root of the penalized residual sum of
  # squares divided by n, the number of observations, evaluated as:
  n <- length(resid(modelglmer))
  return(  sqrt( sum(c(resid(modelglmer),modelglmer@u) ^2) / n ) ) 
} 

dispersion_glmer(model4_fit)
```

## Figure 3.2

Again, using the `emmeans` package.

```{r first logistic plot}
library(emmeans)
glmer_predicted <- emmeans(model4_fit, c("party", "pactotal"), 
                           at = list(party = c("Republican", "Democrat"), 
                                     pactotal = 1:120),
                           type = "response") %>% as.data.frame()

glmer_predicted %>% 
  ggplot(aes(y = prob, x = pactotal, group = party)) +
  geom_line(aes(linetype = party)) +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL), alpha = .05) +
  lims(y = c(0, 1)) +
  labs(x = "Pac Contributions ($K)",
       y = "Predicted Pro-Tobacco Voting Probabilities") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = c(0.9, 0.15),
        legend.title = element_blank())
```

## Table 3.3

This may take several minutes, depending upon the speed of your computer.

```{r final glm fit, message=FALSE, warning=FALSE}
model5_fit <- glmer(vote ~ party + pactotal + acres + 
                      (party + pactotal | legislator) + (1 | state), 
                   data = tobacco_votes, 
                   family = binomial)

summary(model5_fit)
```

## Figure 3.3

Same general procedure as above

```{r second logistic plot}
glmer2_predicted <- emmeans(model5_fit, c("party", "pactotal", "acres"), 
                           at = list(party = c("Republican", "Democrat"), 
                                     pactotal = 1:120,
                                     acres = c(0, 10, 100)),
                           type = "response") %>% as.data.frame()

glmer2_predicted %>% 
  unite("party_acres", party, acres) %>% 
  ggplot(aes(y = prob, x = pactotal, group = party_acres)) +
  geom_line(aes(linetype = party_acres)) +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL), alpha = .05) +
  lims(y = c(0, 1)) +
  labs(x = "Pac Contributions ($K)",
       y = "Predicted Pro-Tobacco Voting Probabilities") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = c(0.85, 0.25),
        legend.title = element_blank())
```

---

